Measurement Bias from Address Aliasing

### 2014-05-27 paper.tex
More fixes to the abstract. Fixed some typos elsewhere.

%The perfomance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
%It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
%While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
%Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
Recent research has shown that there can be significant performance differences based seemingly irrelevant factors like user name length.
In this paper we show an important underlying mechanism resulting in this type of measurement bias on modern Intel microarchitectures.
Our approach is to search for correlations between a large set of hardware performance counters and performance figures from multiple environment contexts.
We find that \emph{address aliasing} can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.

This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
Further, we show that the allocation logic of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
The performance impact of unfavorable memory layout combinations can be significant, with as much as $2x$ speedup in one of our examples.
Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.

@@

\section{Introduction}
Accurately measuring the performance of computer programs is important in order to evaluate new algorithms or compare different implementations.
This process is complicated by \emph{measurment bias}, which can be a challenge for performance analysist and systems researchers.
Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
The reason this happens is because certain variables has the potential to offset memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to things like system environment variables ultimately affect program execution.

@@

By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
To analyse performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-neglegible amount of work.
The overhead can be masked by repeatedly invoking the convolution kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        conv(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
\[
t_{\text{estimate}} = \frac{t_{k} - t_{1}}{k - 1}
\]
Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of perf.

@@

It is worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
The L1 hit rate remains stable across all offsets, and only a neglible number of memory loads actually misses L1.
On the other hand, we see a fairly strong correlation between cycle count and outstanding offcore requests. 
These events count the number of outstanding loads to memory outside the processor core each cycle.
Since we do not see any significant number of loads missing L1, the correlation to outstanding loads is probably a result of stalling and more cycles executed.

@@

\paragraph{Use a special purpose allocator}
We found that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
This goes somewhat agains conventional wisdom that more alignment is better.
The Intel optimization manual actually mentiones this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.

@@

In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing.
Analyzing an example with bias to enviornment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
We show how aliasing can have significant performance impact in practice for algorithms operating on heap allocated buffers, analyzing and example with more than 50 \% performance variation between different memory layouts.
Bacause typical heap allocators will return aliased pointers for large allocations, worst case performance is likely to be the default.

@@

% Acknowledgements
\begin{acks}
This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurment bias on the Intel Ivy Bridge architecture~\cite{MasterThesis}.
\end{acks}

>>>>

%The perfomance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
%It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
%While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
%Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
Recent research has shown that there can be significant performance differences based on seemingly irrelevant factors like user name length.
In this paper we identify an important underlying mechanism that cause this type of measurement bias on modern Intel microarchitectures.
Our approach is to search for correlations between a large set of hardware performance counters and performance figures from multiple environment contexts.
We find that \emph{address aliasing} can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.

This result not only makes accounting for and avoiding measurement bias much simpler, but also enables explicit optimizations for it.
We demonstrate this by implementing runtime detection and correction of address aliasing, eliminating the bias.
Further, we show that the allocation logic of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
The performance impact of unfavorable memory layout combinations can be significant, with as much as $2x$ speedup in one of our examples.
%Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.
In the longer term, our results can enable novel compiler and architectural optimization strategies.

@@

\section{Introduction}
Accurately measuring the performance of computer programs is important in order to evaluate new algorithms or compare different implementations.
This process is complicated by \emph{measurement bias}, which can be a challenge for performance analysist and systems researchers.
Changes to external factors of the system, such as link ordering or the contents of environment variables, has been shown to have potentially significant impacts on performance in real applications~\cite{Mytkowicz:2009:WrongData}.
The reason this happens is because certain variables has the potential to offset memory addresses of code or data, which in turn interacts with various low level hardware buffers at runtime.
Because of the complexity of modern processors, it is often hard or even impossible to predict exactly how changes to things like system environment variables ultimately affect program execution.

@@

By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
To analyze performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-negligible amount of work.
The overhead can be masked by repeatedly invoking the convolution kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        conv(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
\[
t_{\text{estimate}} = \frac{t_{k} - t_{1}}{k - 1}
\]
Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of perf.

@@

It is worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
The L1 hit rate remains stable across all offsets, and only a negligible number of memory loads actually misses L1.
On the other hand, we see a fairly strong correlation between cycle count and outstanding offcore requests. 
These events count the number of outstanding loads to memory outside the processor core each cycle.
Since we do not see any significant number of loads missing L1, the correlation to outstanding loads is probably a result of stalling and more cycles executed.

@@

\paragraph{Use a special purpose allocator}
We found that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
This goes somewhat against conventional wisdom that more alignment is better.
The Intel optimization manual actually mentions this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.

@@

In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing.
Analyzing an example with bias to environment size, we determined that collisions between automatic variables and static data resulted in aliasing for certain stack positions.
Aliasing conditions were triggered because variations in environment size offset the virtual addresses of stack allocated variables.
We show how aliasing can have significant performance impact in practice for algorithms operating on heap allocated buffers, analyzing and example with more than 50 \% performance variation between different memory layouts.
Because typical heap allocators will return aliased pointers for large allocations, worst case performance is likely to be the default.

@@

% Acknowledgements
\begin{acks}
This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel Ivy Bridge architecture~\cite{MasterThesis}.
\end{acks}
